---
title: "Lab 11 - Titanic"
author: "Sophie Boyd"
date: "3-20-26"
output: github_document
---

## Load packages and data

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse) 
library(titanic)
library(readxl)
```

```{r read-data}
titanic3 <- read_excel("data/titanic3.xls",
    col_types = c("numeric", "numeric", "text",
        "text", "numeric", "numeric", "numeric",
        "text", "numeric", "text", "text",
        "text", "text", "text"))

data("titanic_train")
data("titanic_test")
```

```{r clean-data}
# Rename columns to lowercase
names(titanic3) <- names(titanic3) %>% tolower()
names(titanic_train) <- names(titanic_train) %>% tolower()
names(titanic_test) <- names(titanic_test) %>% tolower()

# Convert pclass to factor

titanic_train <- titanic_train %>%
  mutate(pclass_ord = factor(pclass, ordered = TRUE, levels = c(3, 2, 1)))

titanic_test <- titanic_test %>%
  mutate(pclass_ord = factor(pclass, ordered = TRUE, levels = c(3, 2, 1)))

titanic3 <- titanic3 %>%
  mutate(pclass_ord = factor(pclass, ordered = TRUE, levels = c(3, 2, 1)))
```

```{r view-data}
glimpse(titanic_test)
summary(titanic_test)

glimpse(titanic_train)
glimpse(titanic_train)

glimpse(titanic3)
summary(titanic3)
```

titanic_test has 418 observations, titanic_train has 891 observations, and titanic3 has 1309 observations. Available variables include the passengers' ID numbers, class (SES proxy), names, sex, ages, number of siblings/spouses on board, number of parents/children on board, ticket numbers, fare amounts, cabin numbers, embarkation locations, lifeboat numbers (for those who survived), body identification numbers (for those who did not survive and were recovered), and homes/destinations. There are many missing values on cabin number. 

# Part 1: Apparent accuracy

## Exercise 1

```{r m_apparent}
m_apparent <- glm(
  survived ~ sex + pclass,
  data = titanic3,
  family = binomial
)

summary(m_apparent)
```

## Exercise 1.2

```{r p_apparent}
p_apparent <- predict(m_apparent, type = "response")

p_apparent
```

## Exercise 1.3

```{r yhat_apparent}
yhat_apparent <- ifelse(p_apparent > 0.5, 1, 0)

yhat_apparent
```

## Exercise 1.4

```{r acc_apparent}
acc_apparent <- mean(yhat_apparent == titanic3$survived, na.rm = TRUE)

acc_apparent
```

## Exercise 1.5

- Apparent accuracy likely overestimates true predictive performance because it is calculated based on all of the available information about an event. The goal of the predictive model is to predict the unknown, and this capability is not put to the test when all of the known information was used to create the model. 

- Testing on the same data you trained on would also likely give misleadingly good results if you were trying to predict the efficacy of a drug for a particular health outcome.

# Part 2: Holding passengers back

## Exercise 2.1

```{r m_split}
m_split <- glm(
  survived ~ sex + pclass,
  data = titanic_train,
  family = binomial
)

summary(m_split)
```

## Exercise 2.2

```{r train-accuracy}
p_train <- predict(m_split, type = "response")
yhat_train <- ifelse(p_train > 0.5, 1, 0)

acc_train <- mean(yhat_train == titanic_train$survived, na.rm = TRUE)
acc_train
```

## Exercise 2.3

```{r test-accuracy}
p_test <- predict(m_split, newdata = titanic_test, type = "response")
yhat_test <- ifelse(p_test > 0.5, 1, 0)

acc_test <- mean(yhat_test == titanic_test$survived, na.rm = TRUE)
acc_test
```


I got "NaN" as my result above because the titanic_test dataset is missing the "survived" variable. Here is my attempt at merging it in and re-running:

```{r test-accuracy2}
titanic3_survived <- titanic3 %>%
  select(survived, ticket, name)

titanic_test_survived <- left_join(titanic_test, titanic3_survived, by = "name")

p_test <- predict(m_split, newdata = titanic_test_survived, type = "response")
yhat_test <- ifelse(p_test > 0.5, 1, 0)

acc_test <- mean(yhat_test == titanic_test_survived$survived, na.rm = TRUE)
acc_test
```


## Exercise 2.4

- acc_train was .79 and acc_test was .76. acc_train was larger and this pattern is typical because acc_train was applied to the data that informed the model, whereas acc_test was the result of testing a predictive model on data that was not involved in the model's creation. 

- A measure of how well the model predicts future passengers is most useful for Lloyd's.

- If acc_test happened to be higher than acc_train, the logic of holdout testing would still apply. The key feature of holdout testing is that it allows for testing on new data that did not inform the creation of the model. 

# Part 3: Cross validation across timelines

## Exercise 3.1

```{r titanic_cv}
set.seed(11)

titanic_cv <- titanic3 %>%
  filter(!is.na(survived), !is.na(sex), !is.na(pclass)) %>%
  mutate(fold = sample(rep(1:10, length.out = n())))

titanic_cv %>% count(fold)
```

## Exercise 3.2

```{r cv_results}
cv_results <- data.frame(fold = sort(unique(titanic_cv$fold)), accuracy = NA_real_) 
for (j in cv_results$fold) {
  train_j <- titanic_cv %>% filter(fold != j)
  test_j <- titanic_cv %>% filter(fold == j)
  
  m_j <- glm(
    formula = survived ~ sex + pclass,
    data = train_j,
    family = binomial
  )
  
  p_j <- predict(m_j, newdata = test_j, type = "response")
  yhat_j <- ifelse(p_j > 0.5, 1, 0)
  
  cv_results$accuracy[cv_results$fold == j] <- mean(yhat_j == test_j$survived, na.rm = TRUE)
}

cv_results %>% arrange(fold)
```

## Exercise 3.3

```{r cv-summary}
cv_mean <- mean(cv_results$accuracy)
cv_sd <- sd(cv_results$accuracy)
cv_min <- min(cv_results$accuracy)
cv_max <- max(cv_results$accuracy)

c(cv_mean = cv_mean, cv_sd = cv_sd, cv_min = cv_min, cv_max = cv_max)
```

## Exercise 3.4

```{r cv-accuracy-plot}
ggplot(cv_results, aes(x = factor(fold), y = accuracy)) +
  geom_col(fill = "steelblue") +
  geom_hline(yintercept = mean(cv_results$accuracy), linetype = "dashed", color = "red") +
  labs(
    title = "10-Fold Cross Validation: Accuracy by Fold",
    x = "Fold",
    y = "Accuracy"
  ) +
  theme_minimal()
```

## Exercise 3.5

- cv_mean is usually lower than acc_apparent because it is the mean of accuracy rates from models that were tested on new data that was not used for training, whereas acc_apparent is an accuracy rate of a model tested on data that was involved in its training. Testing on data that was used to train a model inflates accuracy.

- cv_sd provides information about how much the accuracy rates of the different test models differed from one another. 

- I would report cv_mean because I have more confidence in an accuracy measure drawn from multiple model tests rather than a single model test. 

# Part 4: When the cutoff is a policy decision

## Exercise 4.1

```{r cutoffs}
cutoffs <- c(0.3, 0.5, 0.7)

cutoff_results <- data.frame(
  cutoff = cutoffs,
  accuracy = NA_real_
)

for (i in seq_along(cutoffs)) {
  c0 <- cutoffs[i]
  
  yhat_c <- ifelse(p_test > c0, 1, 0)
  cutoff_results$accuracy[i] <- mean( yhat_c == titanic_test_survived$survived, na.rm = TRUE)
}

cutoff_results
```

## Exercise 4.2

```{r visualize-cutoffs}
cutoffs_fine <- seq(0.1, 0.9, by = 0.05)

cutoff_results_fine <- data.frame(
  cutoff = cutoffs_fine,
  accuracy = NA_real_
)

for (i in seq_along(cutoffs_fine)) {
  c0 <- cutoffs_fine[i]
  yhat_c <- ifelse(p_test > c0, 1, 0)
  cutoff_results_fine$accuracy[i] <- mean(yhat_c == titanic_test_survived$survived, na.rm = TRUE)
}

ggplot(cutoff_results_fine, aes(x = cutoff, y = accuracy)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Test Accuracy Across Probability Cutoffs",
    x = "Cutoff",
    y = "Accuracy"
  ) +
  theme_minimal()
```


## Exercise 4.3

- The cutoff of 0.7 maximized accuracy. 

- In this scenario, overestimating survival rates carries more danger than underestimating survival rates, as an overestimate may lead to inadequate safety preparation. Therefore, a conservative estimate in this scenario would probably be better than a highly accurate estimate (though it happened to be the case here that the most conservative estimate was also the most accurate).

- In a situation where false positives and false negatives had different costs, I would want to know the expected costs of over- vs. underestimating risk to make a more informed decision. 
